---
title: "Starting with with the dplyr, ggplot2 packages, and the pipe operator |>"
author: "Han Olff changed by zita"
date: "2024-08-29"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

# Working with the dplyr and ggplot2 packages

Load the required libraries :

```{r message = FALSE, warning = FALSE}
# restore your library, keep at the start of every script 
renv::restore()

# clear the R environment
rm(list = ls())

# load the required packages
library(tidyverse)  # this loads a group of related packages, including readr, dplyr and ggplot2
```

We will work with an online database of the Schiermonnikoog transect
study, that you will collect additional data for next week in the field.
In this example, we work with measurements of cockles (a bivalve
mollusc) on their width and length. From the database Epibenthos, we
will work with the table
[FactCockles](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1538766002#gid=1538766002).
See the documentation of the different variables in the table
[MetVariables](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1290622213#gid=1290622213)

### Read the datafile from the Google sheets database

To read the data in R, you first need to know the published csv link of
the FactCockles table. In the database, you can find this link in the
table MetTables. It is produced in Google Sheets throught the menu
File/Share/Publish to web and then selecting the table and output as
csv.

Read the FactCockleSize table with read_csv, which reads it as tibble
(formatted dataframe)

```{r}
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")
print(FactCockleSize)
class(FactCockleSize)
names(FactCockleSize)
nrow(FactCockleSize)
```

Plot with ggplot the relation between cockle thickness (thickness_mm, as
x) and length (length_mm), showing each year with a different color, and
add a regression line through all the years

```{r}

# there is a big outlier, likely a data entry mistake, so we remove it with piping 
FactCockleSize |> dplyr::filter(thickness_mm>500)


# here we take out the outlier with dplyr
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> 
  dplyr::filter(CockleObs_ID!=1531, CockleObs_ID!=469) |>
  dplyr::mutate(year=factor(year)) # and making year a factor 

print(FactCockleSize)

#finding the other outlier 
FactCockleSize |> dplyr::filter(thickness_mm>10 & length_mm <5)

# also remove CockleObs_ID 469

# plot the data 
ggplot2::ggplot(data = FactCockleSize, mapping = aes(x = length_mm, y = thickness_mm, color = as.factor(year),  group = as.factor(year))) + geom_point() + geom_smooth(method = "lm", se = T)




```

Further explore the plot with a regression line through all the data

```{r}
ggplot2::ggplot(data = FactCockleSize, mapping = aes(x = length_mm, y = thickness_mm, color = year,  group = as.factor(year))) + 
  geom_point() + 
  geom_smooth(method = "lm")

# calculate a new model

model_lm <- lm((thickness_mm)~(length_mm), data = FactCockleSize)
summary(model_lm)

# equation : thickness_mm = -0.3534 + 0.6694*length_mm
# a cockle of length 20mm has a predicted thickness of:
-0.3534 + 0.6694 * 20

# this regression is significant, the R2 = 0.90, and the P < 0.001 (***)
```

Make same plot but showing a separate regression line per year

```{r}
# color the points by year, but plot one regression line

    
data <- FactCockleSize |> # what data to use
  dplyr::filter(year %in% c(2014, 2015, 2017, 2018, 2020, 2021)) # only use these years

ggplot2::ggplot(data = data, mapping = aes(x = length_mm, y = thickness_mm)) + # use the filtered data
  geom_point(mapping = aes(col = year)) + # the points are having the color based on year 
  geom_smooth(method = "lm") # add the linear model smoother    
  
```

Make a panel plot where with each year is shown as a separate graph

```{r}


data = FactCockleSize |> # what data to use
  dplyr::filter(year %in% c(2014, 2015, 2017, 2018, 2020, 2021)) |> # only use these years 
  dplyr::mutate(year=factor(year)) # and making year a factor
ggplot2::ggplot(data=data, mapping = aes(x = length_mm, y = thickness_mm)) + # what is on the axes
  geom_point(mapping = aes(col = year)) + # the points are having the color based on year 
  geom_smooth(method = "lm") +
  facet_wrap(~ year) # this makes a panel plot with each year as a separate graph
  

# dropping the NA values is antoher way to do this
drop_na(FactCockleSize) |>
  ggplot2::ggplot(mapping = aes(x = length_mm, y = thickness_mm, color = year,  group = as.factor(year))) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  facet_wrap(~ year)


```

We conclude from this that: \* there were two important outliers in the
dataset that were removed after visual inspection \* the regression
between length and width is abou the same for every year, we safely can
use only length as a proxy for the biomass of an individual cockle
